Derive final annotations based on NBAtlas reference

### 2025-08-18 analyses/cell-type-neuroblastoma-04/final-annotation.Rmd
WIP: add dotplot code, some of which is temporary

theme_set(
  theme_bw() +
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom", 
    aspect.ratio = 1
  )
)

@@

# Pull out metadata we'll want later
sample_metadata <- celltype_df |>
  dplyr::select(sample_id, library_id, is_xenograft) |>
  unique()

# We can remove the merged_sce now
rm(merged_sce)
```

@@

```{r fig.width = 16, fig.height = 16}
umap_df |>
  faceted_umap(
    annotation_column = cell_type_assignment,
    celltype_colors = celltype_pal
  )
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = tumor_cell_classification) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = is_xenograft) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1)))
```

@@

This section shows dot plots of marker gene expression for the final annotations.
Specifically, we show the top 5 highest `log2FC` marker genes per NBAtlas cell type for validation.
Marker genes that are present in three or more cell types are not included.
Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.

```{r}
# data frame of top 5 upregulated genes per cell type
top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 5, "up") |>
  # only markers in no more than 2 categories
  # this is a good middle ground to ensure there are at least 2 markers per cell type
  dplyr::add_count(ensembl_gene_id) |>
  dplyr::filter(n < 3) |>
  dplyr::select(-n) 
```

```{r, fig.width = 24, fig.height = 12}

# get total number of cells per final annotation group and set up y_label
total_cells_df <- full_celltype_df |>
  dplyr::count(label, name = "total_cells") |>
  # order so cell types are grouped
  dplyr::inner_join(label_map_df, by = "label" ) |>
  dplyr::mutate(
    label_recoded = as.factor(label_recoded),
    label_recoded = forcats::fct_relevel(label_recoded, levels(label_recoded_factor))
  ) |>
  dplyr::arrange(label_recoded) |>
  dplyr::mutate(y_label = glue::glue("{label} ({total_cells})")) |>
  # rename as expected for dotplot function - label_recoded needs to be the main column
  dplyr::select(-label_recoded) |>
  dplyr::rename(label_recoded = label)

@@

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  # rename this after we've done all the wrangling
  singler_df = full_celltype_df |> dplyr::rename(label_recoded = label),
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = nbatlas_celltype_pal, 
  min_cells = 0
)
```

>>>>

theme_set(
  theme_bw()
)

@@

# Pull out metadata we'll want later
sample_metadata <- celltype_df |>
  dplyr::select(sample_id, library_id, is_xenograft) |>
  unique()
```

@@

```{r fig.width = 16, fig.height = 16}
umap_theme <- list(
  theme(
    axis.text = element_blank(), 
    axis.ticks = element_blank(), 
    legend.position = "bottom", 
    aspect.ratio = 1
  )
)

umap_df |>
  faceted_umap(
    annotation_column = cell_type_assignment,
    celltype_colors = celltype_pal
  ) + 
  umap_theme
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = tumor_cell_classification) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set2") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

```{r fig.width = 8, fig.height = 8}
ggplot(umap_df) +
  aes(x = UMAP1, y = UMAP2, color = is_xenograft) +
  geom_point(alpha = 0.2, size = 0.3) +
  scale_color_brewer(palette = "Set1") +
  guides(color = guide_legend(override.aes = list(size = 2, alpha = 1))) +
  umap_theme
```

@@

This section shows dot plots of marker gene expression for the final annotations.
Specifically, we show the top 5 highest `log2FC` marker genes per NBAtlas cell type for validation.
Marker genes are taken directly from the NBAtlas paper where they were identified with `Seurat::FindMarkers()`.

Marker genes that are present in three or more cell types are not included.
In addition, the upper limit for the color scale has been censored to a maximum of 6.

```{r}
# Temporary to develop code for next section
annotation_df <- readr::read_tsv(params$annotations_tsv) |>
  dplyr::filter(cell_type_assignment != "Unknown") |>
  # recode cell types to match what NBAtlas uses for their marker genes
  dplyr::mutate(
    dotplot_celltype = dplyr::case_when(
      stringr::str_detect(cell_type_assignment, "monocyte") ~ "Monocyte", 
      stringr::str_detect(cell_type_assignment, "cDC") ~ "cDC", 
      stringr::str_detect(cell_type_assignment, "Neuroendocrine") ~ "Neuroendocrine", 
      .default = cell_type_assignment
    )
  )

# data frame of top 5 upregulated genes per cell type
top_nbatlas_markers_df <- subset_nbatlas_markers(nbatlas_markers_df, 5, "up") |>
  # only markers in no more than 2 categories
  # this is a good middle ground to ensure there are at least 2 markers per cell type
  dplyr::add_count(ensembl_gene_id) |>
  dplyr::filter(n < 3) |>
  dplyr::select(-n) 
```

```{r}
# get total number of cells per final annotation group and set up y_label
total_cells_df <- annotation_df |>
  dplyr::count(dotplot_celltype, name = "total_cells") |>
  # remove infrequent cell types
  dplyr::filter(total_cells > 10) |>
  dplyr::mutate(y_label = glue::glue("{dotplot_celltype} ({total_cells})")) |>
  # order by count, but put Unknown at the end
  dplyr::arrange(desc(total_cells)) |> 
  # amazing new dplyr trick; only this row is TRUE, so it gets moved to the last row
  dplyr::arrange(dotplot_celltype == "Unknown") |>
  # the dotplot function expects this column name
  dplyr::rename(label_recoded = dotplot_celltype)

@@

```{r}
# Prepare the expressed_genes vector
# we only care about if that gene is expressed otherwise we won't waste memory and include it
gene_sums <- rowData(merged_sce) |>
  as.data.frame() |>
  dplyr::select(contains("detected")) |>
  as.matrix() |>
  rowSums()
expressed_genes <- names(gene_sums)[gene_sums > 0]
```

```{r, fig.width = 32, fig.height = 16}
generate_dotplot(
  merged_sce = merged_sce,
  markers_df = top_nbatlas_markers_df,
  # rename this after we've done all the wrangling
  celltype_df = annotation_df |> dplyr::rename(label_recoded = dotplot_celltype),
  total_cells_df = total_cells_df,
  expressed_genes = expressed_genes,
  bar_order = nbatlas_bar_order, 
  celltype_palette = celltype_pal, 
  min_cells = 0
) +
  theme(legend.position = "bottom")
```
